---
title: "Coursera Statistical Inference Project part 2"
author: "Ben de Haan"
date: "22 Aug 2015"
output: pdf_document
---

```{r echo=FALSE}
library(ggplot2)
```

# Introduction and methods
This report discusses the R `ToothGrowth` dataset. The dataset is composed of results on the length of odontoblasts in guinea pigs based on different doses of vitamin C intake. The guinea pigs received either 0.5, 1, or 2 mg/day of vitamin C through orange juice (OJ) or asorbic acid (VC).

```{r echo=FALSE}
str(ToothGrowth)
```

The questions that this report aims to answer are:
* What is the 95% confidence interval for odontoblast length for each dose and intake method?
* Is there a difference in odontoblast length between the orange juice (OJ) and asorbic acid (VC) group?
** $H_0$ : there is no significant difference in odontoblast length between OJ and VC group
* Is there a difference in odontoblas length between the groups for each different dose?
** $H_0$ : there is no significant difference in odontoblast length between groups for each dose

Questions will be answered using t-tests and confidence interval calculations by the T-distribution.

To test the difference between groups, the T-test is used. However, this assumes lengths are normally distributed. If the data is not normally distributed, the following solutions are available: 
1. Confidence intervals can be retrieved through bootstrapping.
2. A log transformation of length may solve the abnormality

# Exploratory Data Analysis

## Summary statistics

```{r echo=FALSE}

group <- c("Orange juice (OJ)", "Asorbic acid (VC)", "Total")
summary.stats <- data.frame(group)

oj.mean <- mean(ToothGrowth[ToothGrowth$supp=="OJ",]$len)
vc.mean <- mean(ToothGrowth[ToothGrowth$supp=="VC",]$len)
tot.mean <- mean(ToothGrowth$len)

means <- c(oj.mean, vc.mean, tot.mean)
summary.stats$means <- means

oj.var <- var(ToothGrowth[ToothGrowth$supp=="OJ",]$len)
vc.var <- var(ToothGrowth[ToothGrowth$supp=="VC",]$len)
tot.var <- var(ToothGrowth$len)

oj.sd <- sd(ToothGrowth[ToothGrowth$supp=="OJ",]$len)
vc.sd <- sd(ToothGrowth[ToothGrowth$supp=="VC",]$len)
tot.sd <- sd(ToothGrowth$len)

```

Confidence intervals for the mean were retrieved as follows:
```{r}
# Calculte the errors
oj.error <- qt(0.975, df = 29) * oj.sd / sqrt(30)
vc.error <- qt(0.975, df = 29) * vc.sd / sqrt(30)
tot.error <- qt(0.975, df = 58) * tot.sd / sqrt(60)

# Calculate the confidence intervals
oj.ci <- oj.mean + c(-1,1) * oj.error
vc.ci <- vc.mean + c(-1,1) * vc.error
tot.ci <- tot.mean + c(-1,1) * tot.error

# Add to summary statistics table
cis.lower <- c(oj.ci[1], vc.ci[1], tot.ci[1])
cis.upper <- c(oj.ci[2], vc.ci[2], tot.ci[2])
summary.stats$cis.lower <- cis.lower
summary.stats$cis.upper <- cis.upper
```

```{r echo=FALSE}
variances <- c(oj.var, vc.var, tot.var)
summary.stats$variances <- variances

sds <- c(oj.sd, vc.sd, tot.sd)
summary.stats$sds <- sds

names(summary.stats) <- c("Group", "Mean", "95% CI lower", "95% CI upper", "Variance", "Standard deviation")
summary.stats

```

The groups are of equal size:
```{r}
table(ToothGrowth$supp)
```

As are the doses within groups:
```{r}
table(ToothGrowth[ToothGrowth$supp=="OJ",]$dose)
table(ToothGrowth[ToothGrowth$supp=="VC",]$dose)
```

```{r echo=FALSE}
p <- ggplot(ToothGrowth, aes(factor(supp), len))
p + geom_boxplot(aes(fill = factor(dose))) +
    xlab("Type of supplement") + ylab("Length") +
    ggtitle("Figure 1: Odontoblast length by supplement and dose")
```

Figure 1 suggests some issues that might require further investigation: 
* Orange juice seems to be more effective at a lower dose, but evens out at the higher doses. 
* Higher doses seem to correspond with a higher odontoblast length


## Normality
Odontoblast lengths seem to be approximately normally distributed, with the mean (black line) and median (cyan) quite close together (see figure 2).

```{r echo=FALSE}
qplot(ToothGrowth$len, binwidth = 1, fill=..count.., main = "Figure 2: Histogram of odontoblast length") + 
    scale_fill_gradient("Count", low = "green", high = "red") +
    geom_vline(xintercept=mean(ToothGrowth$len), colour="black") +
    geom_vline(xintercept=median(ToothGrowth$len), colour="cyan")
```

```{r echo=FALSE}
qqnorm(ToothGrowth$len)
```

The Shapiro-Wilk test of normality reveals that we cannot reject $H_0$ that the lengths are normally distributed (p > 0.05).

```{r}
shapiro.test(ToothGrowth$len)
```

# Results

When we compare the two groups we see that p > 0.05. Therefore, we cannot reject $H_0$ that there is no significant difference between the groups.

```{r}
t.test(ToothGrowth[ToothGrowth$supp=="OJ",]$len,
       ToothGrowth[ToothGrowth$supp=="VC",]$len)
```

If we split up and compare between doses, we have different results. 

For a dose of 0.5 mg/day:
```{r}
t.test(ToothGrowth[ToothGrowth$supp=="OJ" & ToothGrowth$dose==0.5,]$len,
       ToothGrowth[ToothGrowth$supp=="VC" & ToothGrowth$dose==0.5,]$len)
```

Since p < 0.05, we can reject $H_0$. This means that there is a difference between the supplement types for a dose of 0.5 mg/day.

For a dose of 1 mg/day:
```{r}
t.test(ToothGrowth[ToothGrowth$supp=="OJ" & ToothGrowth$dose==1,]$len,
       ToothGrowth[ToothGrowth$supp=="VC" & ToothGrowth$dose==1,]$len)
```

Again, p < 0.05, we can reject $H_0$. This means that there is a difference between the supplement types for a dose of 1 mg/day as well.

For a dose of 2 mg/day:
```{r}
t.test(ToothGrowth[ToothGrowth$supp=="OJ" & ToothGrowth$dose==2,]$len,
       ToothGrowth[ToothGrowth$supp=="VC" & ToothGrowth$dose==2,]$len)
```
Here, p > 0.05, so we cannot reject $H_0$. This means that there is no significant difference between the supplement types for a dose of 2 mg/day.


# Discussion
Assuming that odontoblast length is normally distributed, the results show that there is no significant difference between different intake methods for a dose of 2 mg/day. For a dose of 0.5 or 1 mg/day, orange juice was significantly more effective than asorbic acid.
