---
title: "Statistical Inference course project"
author: "Ben de Haan"
date: "22 Aug 2015"
output: pdf_document
---

# Introduction
In this project, I will investigate the exponential distribution in R and compare it with the Central Limit Theorem. 

Firstly, we need to set a seed to ensure reproducibility. Furthermore, for all simulations, lambda is set to 0.2. I will draw 40 values of exponentials for each iteration in the simulation.

```{r}
set.seed(123465)
lambda = 0.2
n = 40 # Values drawn for each iteration 
```

# Simulation
```{r}
means <- c() # Init vector for storing means
variances <- c()
# Simulate 1000 
for (i in 1:1000){
    values <- rexp(n, lambda) # Get n values
    means[i] <- mean(values) # Store the mean for each iteration
    variances[i] <- var(values) # Store the variance for each iteration
}
```

# Data analysis

## Theoretical mean vs. sample mean
The theoretical mean of the exponential distribution is $1 / lambda$. This means that the theoretical mean is:

```{r}
theoretical.mean <- 1 / lambda
theoretical.mean
```

The sample mean can be inferred by taking the mean from the vector of means. Its 95% confidence interval bounds can be derived by taking the 2.5% and 97.5% values from the distribution, which are upper and lower bound, respectively.

```{r}
mean.bounds <- quantile(variances, probs = c(0.025, 0.975))
mean.lowerBound <- mean.bounds[1]
mean.upperBound <- mean.bounds[2]
mean.lowerBound
mean.upperBound
```

## Theoretical variance vs. sample variance
The theoretical standard deviation of the exponential distribution is $1 / lambda$. Since $var = sd^2$, the variance is:

```{r}
theoretical.sd <- 1 / lambda
theoretical.var <- theoretical.sd^2
theoretical.var
```

The sample variance can be inferred by taking the mean from the vector of variances. Its 95% confidence interval bounds can be derived by taking the 2.5% and 97.5% values from the distribution, which are upper and lower bound, respectively.

```{r}
variance.bounds <- quantile(variances, probs = c(0.025, 0.975))
variance.lowerBound <- variance.bounds[1]
variance.upperBound <- variance.bounds[2]
variance.lowerBound
variance.upperBound
```

## Distribution

The means are distributed as follows:
```{r}
hist(means)
```

Variances are distributed as depicted by the following histogram:
```{r}
hist(variances)
```
